---
title: "Fisheries"
author: "Fredrick Boshe"
date: "09/04/2021"
output: 
  github_document: default
  rmarkdown::github_document: default
---
#  Seafood: The Ultimate Food `r emo::ji("fish")`?

As more people are opting against red meat diet for plant based and sea food, this might be a good time to analyze aquatic food sources. The [environmental impact](https://www.greenpeace.org.uk/news/why-meat-is-bad-for-the-environment/#:~:text=It%20causes%20climate%20change,the%20atmosphere%2C%20accelerating%20global%20warming) of industrialized red meat production and [health concerns](https://www.nytimes.com/2019/09/30/health/red-meat-heart-cancer.html) has made sea food a favourable alternative diet also termed as ["pescatarian diet"](https://www.medicalnewstoday.com/articles/323907#:~:text=What%20is%20a%20pescatarian%20diet%3F&text=In%20the%20pescatarian%20diet%2C%20a,and%20fish%20products%20may%20enhance). 

This project looks to answer two thing:

1.  But is sea food environmentally sustainable? 
2.  Does it pack the nutrients to replace red meat diet?



```{r setup, include=TRUE, results='hide', message=FALSE, warning=FALSE}
#Loading packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(ggcorrplot)
library(jsonlite)
library(httr)
devtools::install_github("hadley/emo")

```

We shall be utilizing the fisheries management dataset from the [National Oceanic and Atmostpheric Administration (NOAA)](https://www.fishwatch.gov/resources). Accessing the latest data using their **FishWatch API**.

This is my first time attempting to work with APIs to access data. A learning curve. 

#### Acessing the data
```{r data, include=TRUE, warning=FALSE, message=FALSE}
#Create a function to request the endpoint and receive a dataframe converted from JSON

fish<- function(endpoint) {
  url <- modify_url("https://www.fishwatch.gov", path = endpoint)
  response<-GET(url)
  
  #tracking status codes
  if(http_error(response)){
    print(status_code(response))
    stop("Something went wrong.", call. = FALSE)
  }
  
  if(http_type(response)!="application/json"){
    stop("API did not return json", call. = FALSE)
  }
  
  #Get content as text
  json_text<- content(response, as="text")
  dataframe<-fromJSON(json_text)
  return(dataframe)
  
}

fish.df<-flatten(fish("/api/species?format=json"))

#Remain with variables useful in the analysis
fish.df<-fish.df%>%
  select(2,7,8,12,14,17:19,26,29,32,33,36,41,45:47,
        49,51,56)
var_name<-data.frame(names(fish.df))
var_name<-var_name%>%
  rename(Variables=names.fish.df.)

  
```

We have reduced it to the essential variables we shall use to evaluate some of the sea food being monitored by NOAA. 


```{r clean, einclude=TRUE, warning=FALSE, message=FALSE}

#The data contains html tags throughout the columns. Clean them out
cleanFun <- function(htmlString) {
  return(gsub("<.*?>", "", htmlString))
}
fish.df<-data.frame(map(fish.df, cleanFun))


```


