---
title: "Fisheries"
author: "Fredrick Boshe"
date: "09/04/2021"
output: 
  github_document: default
  rmarkdown::github_document: default
---
#  Seafood: The Ultimate Food `r emo::ji("fish")`?

As more people are opting against red meat diet for plant based and sea food, this might be a good time to analyze aquatic food sources. The [environmental impact](https://www.greenpeace.org.uk/news/why-meat-is-bad-for-the-environment/#:~:text=It%20causes%20climate%20change,the%20atmosphere%2C%20accelerating%20global%20warming) of industrialized red meat production and [health concerns](https://www.nytimes.com/2019/09/30/health/red-meat-heart-cancer.html) has made sea food a favourable alternative diet also termed as ["pescatarian diet"](https://www.medicalnewstoday.com/articles/323907#:~:text=What%20is%20a%20pescatarian%20diet%3F&text=In%20the%20pescatarian%20diet%2C%20a,and%20fish%20products%20may%20enhance). 

This project looks to answer two thing:

1.  But is sea food environmentally sustainable? 
2.  Does it pack the nutrients to replace red meat diet?



```{r setup, include=FALSE, results='hide', message=FALSE, warning=FALSE}
#Loading packages
library(tidyverse)
library(ggplot2)
library(dplyr)
library(corrplot)
library(plotly)
library(ggcorrplot)
library(jsonlite)
library(httr)
devtools::install_github("hadley/emo")

```

We shall be utilizing the fisheries management dataset from the [National Oceanic and Atmostpheric Administration (NOAA)](https://www.fishwatch.gov/resources). Accessing the latest data using their **FishWatch API**.

This is my first time attempting to work with APIs to access data. A learning curve. 

#### Acessing the data
```{r data, include=TRUE, warning=FALSE, message=FALSE}
#Create a function to request the endpoint and receive a dataframe converted from JSON

fish<- function(endpoint) {
  url <- modify_url("https://www.fishwatch.gov", path = endpoint)
  response<-GET(url)
  
  #tracking status codes
  if(http_error(response)){
    print(status_code(response))
    stop("Something went wrong.", call. = FALSE)
  }
  
  if(http_type(response)!="application/json"){
    stop("API did not return json", call. = FALSE)
  }
  #Get content as text
  json_text<- content(response, as="text")
  dataframe<-fromJSON(json_text)
  return(dataframe)
  
}

fish.df<-as_tibble(flatten(fish("/api/species?format=json")))

#Remain with variables useful in the analysis
fish.df<-fish.df%>%
  select(2,7,8,12,14,17:19,26,29,32,33,36,41,45:47,
        49,51,56)
var_name<-data.frame(names(fish.df))
var_name<-var_name%>%
  rename(Variables=names.fish.df.)

  
```

We have reduced it to the essential variables we shall use to evaluate some of the sea food being monitored by NOAA. 


```{r clean, include=TRUE, warning=FALSE, message=FALSE}
#The data contains html tags throughout the columns. Clean them out
cleanFun <- function(htmlString) {
  return(gsub("<.*?>", "", htmlString))
}
fish.df<-data.frame(map(fish.df, cleanFun))


```


### Fishing and impacts
```{r pop, include=TRUE, warning=FALSE, message=FALSE}
#Categorize population(naive categorization using the first instance of the words "Above", "Unknown" and "Below")
fish.df<-fish.df%>%
  mutate(population=ifelse(str_detect(Population, "[Aa]bove"), "Above target", 
                    ifelse(str_detect(Population, "[Uu]nknown"), "Unknown",
                           ifelse(str_detect(Population, "[Bb]elow|Near"),"Below target", NA))))    


```

```{r df1, echo=FALSE}
Fish_Population<-data.frame(table(fish.df$pop))
Fish_Population<-Fish_Population%>%
  rename(Population=Var1)%>%
  rename(Species=Freq)
Fish_Population

```


```{r availability, include=TRUE, warning=FALSE, message=FALSE}
#Categorize their availability (Annual or seasonal)
fish.df<-fish.df%>%
  mutate(availability=ifelse(str_detect(Availability, "[Yy]ear-?round"), "Year round", "Seasonal"))    

```

```{r df2, echo=FALSE}

Fish_Availability<-data.frame(table(fish.df$availability))
Fish_Availability<-Fish_Availability%>%
  rename(Availability=Var1)%>%
  rename(Species=Freq)
Fish_Availability

```

For fish that are caught fresh in long periods of the year i.e. 6-8 months of the year, and are canned or frozen year long, are coded as available year-round. 


Fish that are caught on strict seasonal basis e.g. June to October, together with fish that are sporadically caught i.e. Wreckfish, are categorized as seasonal.


```{r rate, include=TRUE, warning=FALSE, message=FALSE}
#Categorize the fishing rates
##regex to capture overfishing
rate_regex="i?(closed|prohibited|[Rr]educe(d)?|quota|(rates))"
fish.df<-fish.df%>%
  mutate(fishing_rate=ifelse(str_detect(Fishing.Rate, "[Rr]ecommended|Not"), "Stable",
                             ifelse(str_detect(Fishing.Rate,rate_regex),"Over", "Unknown"))) 

```

```{r df3, echo=FALSE}
Fish_rate<-data.frame(table(fish.df$fishing_rate))
Fish_rate<-Fish_rate%>%
  rename("Fishing rate"=Var1)%>%
  rename(Species=Freq)
Fish_rate

```

How the fishing rates were coded:


1.  **Over**: This include species are experiencing a mandated reduced fishing, species that are in regions where authorities have implemented quotas or rates to protect populations or fishing is out-right closed. e.g Sardines
2.  **Stable**: This include species that that are being fished at "recommended levels" within the fishing regions during the compilation of this data. e.g. Lobster
3.  **Unknown**: This single observation is for the **California Market Squid** whose fishing rate and populations have not be estimated by NOAA.


```{r Env, include=TRUE, warning=FALSE, message=FALSE}
#Categorize the environmental impact of species
fish.df<-fish.df%>%
  mutate(env_effects=ifelse(str_detect(Environmental.Effects, "benefits"), "Net benefit",
                            ifelse(str_detect(Environmental.Effects, "state"), "Federal monitoring",
                                   "Unknown"))) 

```

```{r df4, echo=FALSE}

Fish_env<-data.frame(table(fish.df$env_effects))
Fish_env<-Fish_env%>%
  rename("Environmental Effects"= Var1)%>%
  rename(Species=Freq)
Fish_env

```

Unfortunately, not all species have their environmental effects recorded. Oysters, Mussels, Clams and Geoducks have been quoted to have a "**net benefit** on the environment as they remove excess nutrients and improve water quality". Sugar kelp have the same benefits with the additional carbon fixing where they turn carbon-dioxide into oxygen. 

The Atlantic Salmon and Sablefish are being actively monitored by the federal and state authorities to ensure their fishing have minimal impact on the environment. 


###  Fish Nutrition
```{r nutrition, include=TRUE, warning=FALSE, message=FALSE}
#Convert the columns to numeric and rename to carry their respective units
#Calories column to numeric 
fish.df$Calories<-parse_number(fish.df$Calories) 

#Carbohydrates
fish.df$Carbohydrate<-parse_number(fish.df$Carbohydrate) 
fish.df<-fish.df%>%
  rename("carbs (g/ser)" = Carbohydrate)

#Cholesterol
fish.df$Cholesterol<-parse_number(fish.df$Cholesterol) 
fish.df<-fish.df%>%
  rename("cholesterol (mg/ser)" = Cholesterol)

#Fat
fish.df$Fat..Total<-parse_number(fish.df$Fat..Total) 
fish.df<-fish.df%>%
  rename("fat (g/ser)" = Fat..Total)

#Fiber
##This column has one value in mg and the rest are in g
unique(fish.df$Fiber..Total.Dietary)
fish.df$Fiber..Total.Dietary<-parse_number(fish.df$Fiber..Total.Dietary)

##convert the mg value to g
fish.df$Fiber..Total.Dietary<-ifelse(fish.df$Fiber..Total.Dietary==34, 34/1000, fish.df$Fiber..Total.Dietary)
fish.df$Fiber..Total.Dietary<-round(fish.df$Fiber..Total.Dietary, 2)

fish.df<-fish.df%>%
  rename("fiber (g/ser)" = Fiber..Total.Dietary)

#Protein
fish.df$Protein <-parse_number(fish.df$Protein) 
fish.df<-fish.df%>%
  rename("protein (g/ser)" = Protein)

#Saturated fat
fish.df$Saturated.Fatty.Acids..Total <-parse_number(fish.df$Saturated.Fatty.Acids..Total )
fish.df$Saturated.Fatty.Acids..Total<-round(fish.df$Saturated.Fatty.Acids..Total, 2)
fish.df<-fish.df%>%
  rename("saturated fatty acids (g/ser)" = Saturated.Fatty.Acids..Total)

#Selenium
##Some values are in percentage of daily intake and some in mcg
unique(fish.df$Selenium)
fish.df$Selenium <-parse_number(fish.df$Selenium) 
fish.df$Selenium<-ifelse(fish.df$Selenium==160, 55*1.6, fish.df$Selenium) #55 mcg is the USA daily intake for adults and children above 14.

fish.df<-fish.df%>%
  rename("selenium (mcg/ser)" = Selenium)

#Sodium
fish.df$Sodium <-parse_number(fish.df$Sodium) 
fish.df<-fish.df%>%
  rename("sodium (mg/ser)" = Sodium)

#Sugars
fish.df$Sugars..Total <-parse_number(fish.df$Sugars..Total) 
fish.df<-fish.df%>%
  rename("sugars (g/ser)" = Sugars..Total)

#Serving Weight
fish.df$Serving.Weight <-parse_number(fish.df$Serving.Weight) 
fish.df<-fish.df%>%
  rename("serving weights (g)" = Serving.Weight)


```

```{r mutate, include=FALSE}

#remove redundant columns
fish.df<-fish.df%>%
  select(-Habitat, -Population, -Availability, -Environmental.Effects)


#lower case column names
fish.df<-fish.df%>%
  rename_with(str_to_lower)


```


### Region with the most fish species

```{r region, include=TRUE, fig.height= 5, fig.width= 10, results="hide" }
fig1<-fish.df%>%ggplot(aes(x=noaa.fisheries.region, fill=noaa.fisheries.region))+
  geom_bar()+
  theme(plot.title = element_text(color="black", size=14, face="bold",hjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank(),
        axis.title.x = element_text(color="black", size=12, face="bold"),
        axis.title.y = element_text(color="black", size=12, face="bold"))+
  scale_y_continuous(expand = c(0, 0))+
  labs(title = "Distribution of fish species within \nNOAA fisheries regions",
       x="Region",
       y="No of Species")

mrg <- list(l = 10, r = 40,
          b = 15, t = 100,
          pad = 0)

fig1<-ggplotly(fig1)%>%layout(margin=mrg)

```

```{r fig2, include=TRUE, fig.height= 5, fig.width= 10, results="hide", fig.align="center" }
#Plotly to make the chart more interactive
fig2<- ggplotly(fig1)%>% add_annotations( text="NOAA Fishery Region", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.9, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
  layout( legend=list(y=0.9, yanchor="top" ) )
fig2

```


##  How nutritious is the seafood?

```{r seafood, include=TRUE, warning=FALSE, message=FALSE}
#Using the RDA information from US Dietary guidelines. Some are analyzed as a percentage of calories while others are analyzed in absolute amounts.

#a) Dietary guidelines as a percentage of food calories
fish.df<-fish.df%>%
  mutate(fat_RDA=`fat (g/ser)`*100/93,
         s.fat_RDA=`saturated fatty acids (g/ser)`*100/13,
         carbs_RDA=`carbs (g/ser)`*100/130,
         sugars_RDA=`sugars (g/ser)`*100/25,
         protein_RDA=`protein (g/ser)`*100/56)

#b) Dietary guidelines as in absolute amounts
fish.df<-fish.df%>%
  mutate(cholesterol_RDA=`cholesterol (mg/ser)`*100/300,
         fiber_RDA=`fiber (g/ser)`*100/30,
         sodium_RDA=`sodium (mg/ser)`*100/1300,
         selenium_RDA=`selenium (mcg/ser)`*100/55)

fish.df<-fish.df %>% 
  mutate_at(vars(ends_with("RDA")), funs(round(., 2)))


```



```{r fig3, include=TRUE, fig.height= 8, fig.width= 10, fig.align="center"}

fish_longer<-fish.df%>%
  pivot_longer(cols = c(fat_RDA, s.fat_RDA, carbs_RDA, sugars_RDA, protein_RDA,
                        cholesterol_RDA, fiber_RDA, sodium_RDA, selenium_RDA),
               names_to="nutrient",
               values_to="percent")


#Trend of fish calories and their nutrients
nut.labs<-c("Fat RDA", "Saturated Fatty Acids RDA", "Carbohydrate RDA", "Sugars RDA",
            "Protein RDA", "Cholesterol RDA", "Dietary Fiber RDA", "Sodium RDA", 
            "Selenium RDA")
names(nut.labs)<-c("fat_RDA", "s.fat_RDA", "carbs_RDA", "sugars_RDA", "protein_RDA",
                    "cholesterol_RDA", "fiber_RDA", "sodium_RDA", "selenium_RDA")


fig3<-fish_longer%>%ggplot(aes(x=nutrient, y=percent, color=nutrient))+
  geom_boxplot()+
  theme(plot.title = element_text(color="black", size=14, face="bold",hjust = 0.5),
        axis.text.x = element_blank(),
        axis.ticks = element_blank(),
        legend.title = element_blank(),
        axis.title.x = element_text(color="black", size=12),
        axis.title.y = element_text(color="black", size=12))+
  labs(title="Fish calories and nutritional values \n(percentage RDA)",
       x= "Nutrient",
       y= "Percentage RDA")+
  scale_color_discrete(labels = c("Fat RDA", "Saturated Fatty Acids RDA", "Carbohydrate RDA",
                                 "Sugars RDA","Protein RDA", "Cholesterol RDA", 
                                 "Dietary Fiber RDA", "Sodium RDA", "Selenium RDA"))
fig3



fig4<-plot_ly(data = fish_longer, x=~nutrient, y =~percent, color =~nutrient, 
        type = "box") %>% 
  layout(title="Fish calories and nutritional values \n(percentage RDA)",
         xaxis = list(title='Nutrient',
                      ticktext = list("Fat RDA", "Saturated Fatty Acids RDA", "Carbohydrate RDA",
                                      "Sugars RDA","Protein RDA", "Cholesterol RDA", 
                                      "Dietary Fiber RDA", "Sodium RDA", "Selenium RDA"), 
      tickvals = list("fat_RDA", "s.fat_RDA", "carbs_RDA", "sugars_RDA", "protein_RDA",
                      "cholesterol_RDA", "fiber_RDA", "sodium_RDA", "selenium_RDA"), 
      yaxis = list(title="Percentage RDA")))
fig4


```


```{r nut, include=TRUE, fig.height= 7, fig.width= 10, fig.align="center" }
#Plotly to make the chart more interactive
fig3<- ggplotly(fig3)%>% add_annotations( text="Nutrition", xref="paper", yref="paper",
                  x=1.02, xanchor="left",
                  y=0.9, yanchor="bottom",    # Same y as legend below
                  legendtitle=TRUE, showarrow=FALSE ) %>%
  layout(legend=list(y=0.9, yanchor="top"),
          margin=mrg)
fig3

```